---
title: "Calibration under interventions: rticausal and ipeval"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration under interventions: rticausal and ipeval}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 5
)
```

This article reproduces the point-intervention calibration example from the
[`ipeval` vignette](https://cran.r-project.org/web/packages/ipeval/vignettes/ipeval.html)
and compares the resulting calibration curves with `rticausal`.

Both workflows evaluate predictions under a hypothetical intervention. `ipeval` estimates treatment weights internally from a supplied treatment model, whereas `rticausal` intentionally presents this through an API style consistent with the `rtichoke` ecosystem, consuming caller-supplied inverse-probability weights.

## Example data and prediction models

```{r data-models}
library(ipeval)
library(rticausal)

simulate_data <- function(n, seed) {
  set.seed(seed)
  data <- data.frame(id = seq_len(n))
  data$L <- rnorm(n)
  data$A <- rbinom(n, 1, plogis(2 * data$L))
  data$P <- rnorm(n)
  data$Y <- rbinom(
    n,
    1,
    plogis(0.5 + data$L + 1.25 * data$P - 0.9 * data$A)
  )
  data
}

df_dev <- simulate_data(n = 5000, seed = 1)
df_val <- simulate_data(n = 10000, seed = 2)

model_naive <- glm(Y ~ A + P, family = "binomial", data = df_dev)

trt_model_dev <- glm(A ~ L, family = "binomial", data = df_dev)
propensity_dev <- predict(trt_model_dev, type = "response")
iptw_dev <- 1 / ifelse(df_dev$A == 1, propensity_dev, 1 - propensity_dev)
model_causal <- glm(
  Y ~ A + P,
  family = "binomial",
  data = df_dev,
  weights = iptw_dev
)
```

## Calibration if nobody were treated: A = 0

In this section, we evaluate the counterfactual calibration if everyone received treatment $A = 0$.

::: {.panel-tabset group="package"}

## ipeval

In `ipeval`, intervention-specific calibration is computed via `ip_score()`, which fits the treatment model internally to estimate weights.

```{r ipeval-0, fig.width=5, fig.height=5}
score_0 <- ip_score(
  object = list(
    "naive" = model_naive,
    "causal" = model_causal
  ),
  data = df_val,
  outcome = Y,
  treatment_formula = A ~ L,
  treatment_of_interest = 0,
  metrics = "calplot",
  null_model = FALSE,
  quiet = TRUE
)

plot(score_0, main = "Calibration if everyone had A = 0")
```

## rticausal

In `rticausal`, predictions under intervention and inverse-probability weights are supplied explicitly by the caller, maintaining alignment with the `rtichoke` ecosystem API style.

```{r rticausal-0}
trt_model_val <- glm(A ~ L, family = "binomial", data = df_val)
propensity_val <- predict(trt_model_val, type = "response")

pred_0 <- list(
  "naive" = predict(
    model_naive,
    newdata = transform(df_val, A = 0),
    type = "response"
  ),
  "causal" = predict(
    model_causal,
    newdata = transform(df_val, A = 0),
    type = "response"
  )
)
weights_0 <- 1 / (1 - propensity_val)

create_calibration_curve(
  probs = pred_0,
  reals = df_val$Y,
  treats = df_val$A,
  intervention = 0,
  weights = weights_0,
  groups = 8,
  size = 500
)
```

:::

## Calibration if everybody were treated: A = 1

Here, we evaluate counterfactual calibration if everyone received treatment $A = 1$.

::: {.panel-tabset group="package"}

## ipeval

```{r ipeval-1, fig.width=5, fig.height=5}
score_1 <- ip_score(
  object = list(
    "naive" = model_naive,
    "causal" = model_causal
  ),
  data = df_val,
  outcome = Y,
  treatment_formula = A ~ L,
  treatment_of_interest = 1,
  metrics = "calplot",
  null_model = FALSE,
  quiet = TRUE
)

plot(score_1, main = "Calibration if everyone had A = 1")
```

## rticausal

```{r rticausal-1}
trt_model_val <- glm(A ~ L, family = "binomial", data = df_val)
propensity_val <- predict(trt_model_val, type = "response")

pred_1 <- list(
  "naive" = predict(
    model_naive,
    newdata = transform(df_val, A = 1),
    type = "response"
  ),
  "causal" = predict(
    model_causal,
    newdata = transform(df_val, A = 1),
    type = "response"
  )
)
weights_1 <- 1 / propensity_val

create_calibration_curve(
  probs = pred_1,
  reals = df_val$Y,
  treats = df_val$A,
  intervention = 1,
  weights = weights_1,
  groups = 8,
  size = 500
)
```

:::

## Comparison and interpretation

`rticausal`'s intervention calibration path intentionally reproduces `ipeval`'s calibration grouping algorithm so that setting `groups = 8` yields equivalent calibration coordinates. At the same time, `rticausal` maintains the `rtichoke` visual layout (including predicted probability distributions) and requires caller-supplied weights, cleanly decoupling weight estimation from performance evaluation.
